Management induced changes of soil organic carbon on global croplands

### 2020-12-17 SOCbudgetPaper/SOCbudget.tex
update (fix citations, affilations, figure position, LPJmL4Paper to LPJmL4, sidewaytable)

\Author[true]{Kristine}{Karstens}
\Author[1]{Benjamin Leon}{Bodirsky}
\Author[1]{Jan Philipp}{Dietrich}
\Author[2]{Marta}{Dondini}
\Author[1]{Jens}{Heinke}
\Author[2]{Matthias}{Kuhnert}
\Author[1]{Christoph}{Müller}
\Author[1]{Susanne}{Rolinski}
\Author[2]{Pete}{Smith}
\Author[1]{Isabelle}{Weindl}
\Author[1]{Hermann}{Lotze-Campen}
\Author[1]{Alexander}{Popp}

@@

Soil Organic Carbon (SOC), the amount of organic carbon stored in the Earth's soil, is the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times (\citep{batjes_total_1996}). Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO\textsubscript{2} in the atmosphere (\citep{friedlingstein_global_2019}, \citep{minasny_soil_2017}). The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (\citep{batjes_total_1996}, \citep{sanderman_soil_2017}).

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content (\citep{batjes_harmonized_2016}, \citep{hengl_soilgrids250m_2017}, \citep{fao_global_2018}). However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological drivers, they alter terrestrial carbon pools over much shorter time scales and are currently one of most dominant drivers of SOC changes on managed land.

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions), which is the amount of organic carbon soils have lost under cultivation compared to a potential natural vegetated state. \citep{sanderman_soil_2017} identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter (\citep{lal_world_2001}).

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions (\citep{friedlingstein_global_2019}). While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest (\citep{strassmann_simulating_2008}, \citep{betts_climate_2015}). BKMs are designed to estimate LUC related emissons but ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization, and agricultural management (\citep{friedlingstein_global_2019}, \citep{houghton_carbon_2012}, \citep{hansis_relevance_2015}).

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance (e.g.~\citep{bondeau_modelling_2007}, \citep{lindeskog_implications_2013}). \citep{pugh_simulated_2015} explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historic management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historic LUC and the slow decline of SOC under cropland before it reaches a new equilibrium.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g.~growing seasons (\citep{portmann_mirca2000global_2010}) multiple cropping systems (\citep{waha_multiple_2020}), irrigation systems (\citep{jagermeyr_water_2015})) or as stylized scenarios (e.g.~\citep{pugh_simulated_2015}, \citep{lutz_simulating_2019}).

More data sets on spatially explicit agricultural management time series with global coverage become available (e.g.~on tillage systems, see \citep{porwollik_generating_2018}, \citep{prestele_spatially_2018}) and model approaches are increasingly developed to project the dynamics of management systems into the future (e.g.~\citep{iizumi_modeling_2019}, \citep{minoli_modelling_2019}), but have -- to our knowledge -- not found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

Field-scale models (\citet{del_grosso_simulated_2001}{]}, \citep{coleman_simulating_1997},\citep{smith_estimating_2010}, \citep{taghizadeh-toosi_c-tool_2014}) are able to better account for historic agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge.

@@

In Sect. \ref{sec:carbonbudget} we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}). We additionally describe how we configured and extended the steady-state method (for model code see \citep{karstens_mrsoil_2020}). In Sect. \ref{sec:tier1} we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines (\citep{eggleston_ipcc_2006}, \citep{calvo_buendia_ipcc_2019}).
Section \ref{sec:agrimanagement} provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices (for model code see \citep{bodirsky_mrcommons_2020}). In Sect. \ref{sec:scenarios} we define the scenarios used to complement our historic model results.

@@

Following the Tier 2 steady-state approach of the refinement of the IPCC guidelines vol.~4 (\citep{calvo_buendia_ipcc_2019}; referred to as \textit{steady-state method} in the following), we estimate soil organic carbon (\(SOC\)) stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current \(SOC\) state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period:
(1) We calculate annual land-use type-specific steady states and decay rates for \(SOC\) stocks (Sect. \ref{sec:steadystates});
(2) we account for land conversion by transferring \(SOC\) from and to natural vegetation (Sect. \ref{sec:carbontransfer}),
(3) we estimate \(SOC\) stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \ref{sec:totalsoc}).
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \ref{sec:initsoc}).

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the steady-state method for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model (\citep{parton_analysis_1987}). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below- and above-ground residues left or returned to the field (see Sect. \ref{sec:residues}) and manure inputs (see Sect. \ref{sec:livstmanure}); for natural vegetation litterfall (including fine root turnover) (\citep{schaphoff_lpjml4_2018}) is the only source of carbon inflow to the soil. Following the IPCC guidelines \citep{calvo_buendia_ipcc_2019}, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective \(SOC\) sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \ref{tab:datasourceinputs}.

  \begin{table*}[h]
  \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
  \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.20\textwidth} p{0.35\textwidth}}
  \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
  \middlehline
  \multirow{3}{*}{Cropland} & above-ground residues, & \multirow{3}{*}{\begin{minipage}[t]{0.35\columnwidth}\raggedright
                                                                      \cite{faostat_faostat_2016}, \\
                                                                      LPJmL4 \cite{schaphoff_lpjml4_2018},\\
                                                                      \cite{weindl_livestock_2017} \end{minipage}} & 
                                                      \multirow{3}{*}{\begin{minipage}[t]{0.35\columnwidth}\raggedright
                                                                      LG:C generic values according to Table 5.5B,\\
                                                                      5.5C in IPCC \cite{calvo_buendia_ipcc_2019},\\
                                                                      crop-specific N:C from \cite{bodirsky_n2o_2012}
                                                                      \end{minipage}} \\
                            & below-ground residues, &  &  \\
                            & manure                &  &  \\ 
                            \hline
  Natural vegetation        & annual litterfall &  \cite{schaphoff_lpjml4_2018}
                            & \begin{minipage}[t]{0.35\columnwidth}\raggedright\strut 
                                IPCC \cite{calvo_buendia_ipcc_2019} and\\ 
                                CENTURY (\cite{century_model_2000}) \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

@@

For natural vegetation, we assume rainfed and non-tilled conditions, whereas for cropland, we distinguish the effect of different tillage (see Sect. \ref{sec:tillage}) and irrigation (see Sect. \ref{sec:irrigation}) practices on decay rates. We calculated area weighted means for \(till\) and \(wat\) on cropland for each grid cell, using area shares for the different tillage and irrigation pratices. Data sources as well as used parameters for the different decay drivers for all land-use types are listed in Table \ref{tab:datasourcedecay}; equations are displayed by equation 5.0B--5.0F in \citep{calvo_buendia_ipcc_2019}.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm 
                                     & \cite{hengl_soilgrids250m_2017} \\
                      \cline{2-4}

                      & Mircobial activity & air temperature & \cite{harris_version_2020} \\
                      \cline{2-4}

                      & Water restriction & precipitation \& potential evapotranspiration & \cite{harris_version_2020} \\
                      \cline{1-4}
\multirow{2}{*}{\begin{minipage}[t]{0.2\columnwidth}\raggedright\strut Cropland\\(additionally)\strut\end{minipage}} & Water restriction*  & irrigation  & Sect. \ref{sec:irrigation} \\ 
                      \cline{2-4}

@@

According to the IPCC guidelines \(SOC\) changes can be expressed as the difference of two consecutive years (see Eq. 5.0A in \citep{calvo_buendia_ipcc_2019}). This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we defined the absolute and relative SOC changes in relation to a potential natural state \(SOC^{\mathrm{pnv}}\) under the same climatic conditions in grid cell \(i\) at time \(t\), that is based on the natural vegetation \(SOC\) calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes \(\Delta SOC\) and relative changes \(F^{\mathrm{SCF}}\) are thus given by

@@

Note that the absolute changes \(\Delta SOC\) can be also interpreted as the SOC debt (\citep{sanderman_soil_2017}) due to human cropping activities; whereas relative changes \(F^{\mathrm{SCF}}\) can be considered stock change factors as defined within the IPCC guidelines of 2006 \citep{eggleston_ipcc_2006}. Moreover, \(\Delta SOC\) is equivalent to the negated cumulative SOC component of human land-use change emissions (\citep{pugh_simulated_2015}).

@@

Additionally to the steady-state method \citep{calvo_buendia_ipcc_2019} and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines (\citep{eggleston_ipcc_2006}, \citep{calvo_buendia_ipcc_2019}). Here, stocks are calculated via stock change factors (\(F^{\mathrm{SCF}}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of \(F^{\mathrm{SCF}}\) are differentiated by different crops, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, estimates of \(F^{\mathrm{SCF}}\) vary for different climatic zones (\(c\)) specified by the IPCC (see Fig. \ref{fig:CLIMzone}). The actual \(SOC\) stocks are thus calculated based on a given reference stock \(SOC^{\mathrm{ref}}\) by

@@

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis, we use the default \(F^{\mathrm{SCF}}\) from the Tier 1 method of \citep{eggleston_ipcc_2006} and \citep{calvo_buendia_ipcc_2019} as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Land-use patterns are based on the Land-Use Harmonization 2 \citep{hurtt_harmonization_2020} data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data \citep{faostat_faostat_2016} (see FAO2LUH2MAG\_croptypes.csv in \citep{bodirsky_mrcommons_2020}). By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level.
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

Crop residue production and management is based on a revised methodology of \citep{bodirsky_n2o_2012} and key aspects are explained here given its central role for soil carbon modeling. Starting from harvested crop production (\(CP\)) estimates and their respective harvested crop area (\(CA\)), we estimate above-ground (\(AGR\)) and below-ground (\(BGR\)) residual biomass using crop group (\(cg\)) specific harvest index values (\(HI\)) and root:shoot ratios (\(RS\)) as follows

@@

Following the IPCC guidelines, we split the harvest index into a production (\(HI^{\mathrm{prod}}\)) and an area dependent (\(HI^{\mathrm{area}}\)) fraction \citep{eggleston_ipcc_2006}. Deviating from \citep{bodirsky_n2o_2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all \(BGR\) are left in the soil, whereas \(AGR\) can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

A country-specific fixed share of the \(AGR\) is assumed to be burned on field depending on the per-capita income of the country. Following \citep{smil_nitrogen_1999}) we assume a burn share of \(25\%\) for low-income countries according to World Bank definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80--90\% of the carbon in the crop residues burned in the fields is lost within the combustion process \citep{eggleston_ipcc_2006}.

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in \citep{bodirsky_mrcommons_2020}). Residue feed demand for five different livestock groups is based on country-specific feed baskets (see \citep{weindl_livestock_2017}), that differentiate between the residue groups and take available \(AGR\) biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5\% and a fuel-share of 10\% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining \(AGR\) as well as all \(BGR\) are expected to be left on the field. We limit high residue return rates to at most \(10\unit{tC\,ha}^{-1}\) in order to correct for outliers.

@@

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization and parameter choices, we perform a set of scenario runs. In the following sections we outline name and idea of these scenarios (for technical implementation see \citep{karstens_mrsoil_2020}).

@@

Detailed results for the spatially explicit global SOC budget including intermediate results on input data as well as SOC stock results for all scenario runs can be found in \citep{karstens_model_2020}. In the following, the most important results (see \citep{karstens_result_2020} for post-processing script) are summarized.

>>>>

\Author[1, 3]{Kristine}{Karstens}
\Author[1]{Benjamin Leon}{Bodirsky}
\Author[1]{Jan Philipp}{Dietrich}
\Author[2]{Marta}{Dondini}
\Author[1]{Jens}{Heinke}
\Author[2]{Matthias}{Kuhnert}
\Author[1]{Christoph}{Müller}
\Author[1]{Susanne}{Rolinski}
\Author[2]{Pete}{Smith}
\Author[1]{Isabelle}{Weindl}
\Author[1, 3]{Hermann}{Lotze-Campen}
\Author[1]{Alexander}{Popp}

@@

Soil Organic Carbon (SOC), the amount of organic carbon stored in the Earth's soil, is the largest terrestrial organic carbon pool. It exceeds the carbon in the atmospheric and vegetation pools multiple times \citep{batjes_total_1996}. Even small changes in processes affecting SOC lead therefore to substantial shifts in the terrestrial carbon cycle and influence the amount of CO\textsubscript{2} in the atmosphere \citep{friedlingstein_global_2019, minasny_soil_2017}. The specific amount of carbon stored in soils globally is quantified with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile \citep{batjes_total_1996, sanderman_soil_2017}.

Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time. Recent studies have focused on the evaluation of total SOC stocks of the world as well as on the spatial disaggregation of soil properties such as SOC content \citep{batjes_harmonized_2016, hengl_soilgrids250m_2017, fao_global_2018}. However, these studies often do not include human interventions, like land cover change and agricultural management, in their analysis. Compared to climatic and geological drivers, they alter terrestrial carbon pools over much shorter time scales and are currently one of most dominant drivers of SOC changes on managed land.

The anthropogenic impact can be measured by the SOC debt (also referred to as SOC component of land-use change emissions), which is the amount of organic carbon soils have lost under cultivation compared to a potential natural vegetated state. Sanderman et al.~\citeyearpar{sanderman_soil_2017} identified the anthropogenic SOC debt for the first meter of the soil profile due to land cover change at around 116 GtC (37 GtC for the first 30 cm), compared to previous estimates of 60--130 GtC for the first meter \citep{lal_world_2001}.

Global assessments of the carbon cycle via dynamic global vegetation models (DGVMs) and Earth System Models (ESMs) or bookkeeping models (BKMs) have analyzed SOC losses as part of a comprehensive evaluation of the global carbon budget and land-use change (LUC) emissions \citep{friedlingstein_global_2019}. While providing estimates of the magnitude of SOC losses due to land cover change, most models lack a detailed consideration of agricultural management. Earlier DGVM and ESM based assessments only considered changes in land cover, but ignored the removal of biomass at harvest \citep{strassmann_simulating_2008, betts_climate_2015}. BKMs are designed to estimate LUC related emissons but ignore changes in SOC due to climate change, CO\textsubscript{2} fertilization, and agricultural management \citep{friedlingstein_global_2019, houghton_carbon_2012, hansis_relevance_2015}.

Managed agricultural systems were introduced in greater detail to DGVMs and ESMs to improve the assessment of the terrestrial carbon balance \citep[e.g.~][]{bondeau_modelling_2007, lindeskog_implications_2013}. Pugh et al.~\citeyearpar{pugh_simulated_2015} explicitly consider agricultural management in the form of tillage, irrigation and biomass extraction at harvest, but worked with stylized scenarios rather than with historic management data. They also showed the importance of accounting for the land-use history, as many carbon emissions from agricultural soils are caused by historic LUC and the slow decline of SOC under cropland before it reaches a new equilibrium.

In global-scale carbon cycle assessments, management systems are typically represented as spatially explicit patterns that are static in time (e.g.~growing seasons \citep{portmann_mirca2000global_2010} multiple cropping systems \citep{waha_multiple_2020}, irrigation systems \citep{jagermeyr_water_2015}) or as stylized scenarios \citep[e.g.~][\citet{lutz_simulating_2019}]{pugh_simulated_2015}.

More data sets on spatially explicit agricultural management time series with global coverage become available (e.g.~on tillage systems, see \citep{porwollik_generating_2018}, \citep{prestele_spatially_2018}) and model approaches are increasingly developed to project the dynamics of management systems into the future (e.g.~\citep{iizumi_modeling_2019}, \citep{minoli_modelling_2019}), but have --- to our knowledge --- not found their way into comprehensive assessments of the terrestrial carbon cycle in DGVMs and BKMs.

Field-scale models \citep[\citet{coleman_simulating_1997}, \citet{smith_estimating_2010}, \citet{taghizadeh-toosi_c-tool_2014}]{del_grosso_simulated_2001} are able to better account for historic agricultural management if detailed information on crop yield levels, fertilizer inputs and various other on-farm measures is available for the studied sites. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global domain remains a complex challenge.

@@

In Sect. \ref{sec:carbonbudget} we introduce the basic concept of SOC dynamics as applied in this study and explained in more detail within the refinement of the IPCC guidelines vol.~4 \citep{calvo_buendia_ipcc_2019}. We additionally describe how we configured and extended the steady-state method (for model code see \citep{karstens_mrsoil_2020}. In Sect. \ref{sec:tier1} we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines \citep{eggleston_ipcc_2006, calvo_buendia_ipcc_2019}.
Section \ref{sec:agrimanagement} provides a detailed description of the global, gridded management data used to drive the model, including crop production levels, residue input rates, manure amendments, and the adoption of irrigation and tillage practices \citep[for model code see][]{bodirsky_mrcommons_2020}. In Sect. \ref{sec:scenarios} we define the scenarios used to complement our historic model results.

@@

Following the Tier 2 steady-state approach of the refinement of the IPCC guidelines vol.~4 \citep{calvo_buendia_ipcc_2019}; referred to as \textit{steady-state method} in the following), we estimate soil organic carbon (\(SOC\)) stocks and their change over time for cropland at half-degree resolution from 1975 to 2010. We restrict our analysis to the first 0-30 cm of the soil profile. Moreover, we assume the current \(SOC\) state converges towards a steady state, which itself depends on biophysical, climatic and agronomic conditions.
Therefore, we take the following three steps for each year of our simulation period:
(1) We calculate annual land-use type-specific steady states and decay rates for \(SOC\) stocks (Sect. \ref{sec:steadystates});
(2) we account for land conversion by transferring \(SOC\) from and to natural vegetation (Sect. \ref{sec:carbontransfer}),
(3) we estimate \(SOC\) stocks and changes based on the stocks of the previous time step, the steady state stocks and the decay rate (Sect. \ref{sec:totalsoc}).
To initialize the first year of our simulation period we use a spin-up period of 74 years (Sect. \ref{sec:initsoc}).

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{\mathrm{eq}}\) are given by
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being the carbon inputs to the soil, \(k\) denotes the soil organic carbon decay rate. This equation is valid for all grid cells \(i\) and all years \(t\). We use the steady-state method for our calculations, which assumes three soil carbon sub-pools \(sub\) (active, slow and passive) and interactions between them, following the approach in the Century model \citep{parton_analysis_1987}. Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type \(lu\) specific.
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below- and above-ground residues left or returned to the field (see Sect. \ref{sec:residues}) and manure inputs (see Sect. \ref{sec:livstmanure}); for natural vegetation litterfall including fine root turnover \citep{schaphoff_lpjml4_2018} is the only source of carbon inflow to the soil. Following the IPCC guidelines \citep{calvo_buendia_ipcc_2019}, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective \(SOC\) sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \ref{tab:datasourceinputs}.

  \begin{table*}[h]
  \caption{Type and data sources for carbon inputs and parameterization to different land-use types }
  \begin{tabular}{p{0.15\textwidth} p{0.20\textwidth} p{0.20\textwidth} p{0.35\textwidth}}
  \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
  \middlehline
  \multirow{3}{*}{Cropland} & above-ground residues, & \multirow{3}{*}{\begin{minipage}[t]{0.35\columnwidth}\raggedright
                                                                      \citet{faostat_faostat_2016}, \\
                                                                      \citet{schaphoff_lpjml4_2018},\\
                                                                      \citet{weindl_livestock_2017} \end{minipage}} & 
                                                      \multirow{3}{*}{\begin{minipage}[t]{0.35\columnwidth}\raggedright
                                                                      LG:C generic values according to Table 5.5B,\\
                                                                      5.5C from IPCC \citep{calvo_buendia_ipcc_2019},\\
                                                                      crop-specific N:C from \citet{bodirsky_n2o_2012}
                                                                      \end{minipage}} \\
                            & below-ground residues, &  &  \\
                            & manure                &  &  \\ 
                            \hline
  Natural vegetation        & annual litterfall &  \citet{schaphoff_lpjml4_2018}
                            & \begin{minipage}[t]{0.35\columnwidth}\raggedright\strut 
                                IPCC \citep{calvo_buendia_ipcc_2019} and\\ 
                                CENTURY (\citep{century_model_2000}) \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

@@

For natural vegetation, we assume rainfed and non-tilled conditions, whereas for cropland, we distinguish the effect of different tillage (see Sect. \ref{sec:tillage}) and irrigation (see Sect. \ref{sec:irrigation}) practices on decay rates. We calculated area weighted means for \(till\) and \(wat\) on cropland for each grid cell, using area shares for the different tillage and irrigation pratices. Data sources as well as used parameters for the different decay drivers for all land-use types are listed in Table \ref{tab:datasourcedecay}; equations are displayed by equation 5.0B--5.0F in Calvo Buendia et al.~\citeyearpar{calvo_buendia_ipcc_2019}.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm 
                                     & \citet{hengl_soilgrids250m_2017} \\
                      \cline{2-4}

                      & Mircobial activity & air temperature & \citet{harris_version_2020} \\
                      \cline{2-4}

                      & Water restriction & precipitation \& potential evapotranspiration & \citet{harris_version_2020} \\
                      \cline{1-4}
\multirow{2}{*}{\begin{minipage}[t]{0.2\columnwidth}\raggedright\strut Cropland\\(additionally)\strut\end{minipage}} & Water restriction*  & irrigation  & Sect. \ref{sec:irrigation} \\ 
                      \cline{2-4}

@@

According to the IPCC guidelines \(SOC\) changes can be expressed as the difference of two consecutive years \citep[see Eq. 5.0A in][]{calvo_buendia_ipcc_2019}. This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we defined the absolute and relative SOC changes in relation to a potential natural state \(SOC^{\mathrm{pnv}}\) under the same climatic conditions in grid cell \(i\) at time \(t\), that is based on the natural vegetation \(SOC\) calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes \(\Delta SOC\) and relative changes \(F^{\mathrm{SCF}}\) are thus given by

@@

Note that the absolute changes \(\Delta SOC\) can be also interpreted as the SOC debt \citep{sanderman_soil_2017} due to human cropping activities; whereas relative changes \(F^{\mathrm{SCF}}\) can be considered stock change factors as defined within the IPCC guidelines of 2006 \citep{eggleston_ipcc_2006}. Moreover, \(\Delta SOC\) is equivalent to the negated cumulative SOC component of human land-use change emissions \citep{pugh_simulated_2015}.

@@

Additionally to the steady-state method \citep{calvo_buendia_ipcc_2019} and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC Tier 1 approach of IPCC guidelines \citep{eggleston_ipcc_2006, calvo_buendia_ipcc_2019}. Here, stocks are calculated via stock change factors (\(F^{\mathrm{SCF}}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. Estimates of \(F^{\mathrm{SCF}}\) are differentiated by different crops, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, estimates of \(F^{\mathrm{SCF}}\) vary for different climatic zones (\(c\)) specified by the IPCC (see Fig. \ref{fig:CLIMzone}). The actual \(SOC\) stocks are thus calculated based on a given reference stock \(SOC^{\mathrm{ref}}\) by

@@

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis, we use the default \(F^{\mathrm{SCF}}\) from the Tier 1 method of Eggleston et al.~\citeyearpar{eggleston_ipcc_2006} and Calvo Buendia et al.~\citep{calvo_buendia_ipcc_2019} as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Land-use patterns are based on the Land-Use Harmonization 2 \citep{hurtt_harmonization_2020} data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups \citep[see FAO2LUH2MAG\_croptypes.csv in][]{bodirsky_mrcommons_2020}, applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data \citep{faostat_faostat_2016}. By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level.
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

Crop residue production and management is based on a revised methodology of Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012} and key aspects are explained here given its central role for soil carbon modeling. Starting from harvested crop production (\(CP\)) estimates and their respective harvested crop area (\(CA\)), we estimate above-ground (\(AGR\)) and below-ground (\(BGR\)) residual biomass using crop group (\(cg\)) specific harvest index values (\(HI\)) and root:shoot ratios (\(RS\)) as follows

@@

Following the IPCC guidelines, we split the harvest index into a production (\(HI^{\mathrm{prod}}\)) and an area dependent (\(HI^{\mathrm{area}}\)) fraction \citep{eggleston_ipcc_2006}. Deviating from Bodirsky et al.~\citeyearpar{bodirsky_n2o_2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all \(BGR\) are left in the soil, whereas \(AGR\) can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

A country-specific fixed share of the \(AGR\) is assumed to be burned on field depending on the per-capita income of the country. Following Smil \citeyearpar{smil_nitrogen_1999} we assume a burn share of \(25\%\) for low-income countries according to World Bank definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group, 80--90\% of the carbon in the crop residues burned in the fields is lost within the combustion process \citep{eggleston_ipcc_2006}.

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in Bodirsky et al.~\citeyearpar{bodirsky_mrcommons_2020}. Residue feed demand for five different livestock groups is based on country-specific feed baskets \citep[see][]{weindl_livestock_2017}, that differentiate between the residue groups and take available \(AGR\) biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5\% and a fuel-share of 10\% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining \(AGR\) as well as all \(BGR\) are expected to be left on the field. We limit high residue return rates to at most \(10\unit{tC\,ha}^{-1}\) in order to correct for outliers.

@@

To highlight the impact of changing management effects and to assess the sensitivity of the model towards different initialization and parameter choices, we perform a set of scenario runs. In the following sections we outline name and idea of these scenarios \citep[for technical implementation see][]{karstens_mrsoil_2020}.

@@

Detailed results for the spatially explicit global SOC budget including intermediate results on input data as well as SOC stock results for all scenario runs can be found in Karstens \citeyearpar{karstens_model_2020}. In the following, the most important results \citep[see][]{karstens_result_2020} for post-processing script) are summarized.
